---
title: "NC SAT Criterion A Analyses"
Author: "Scott K Anderson"
Date: "2023-12-14"
output: 
  html_document:
    toc: true
    toc_dept: 2
always_allow_hmtl: TRUE
--- 
```{r setup, includ = FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# Introduction
This R script implements the NC Wildlife Resources Commission Species Assessment Tool Criterion A analysis: random re-sampling (Fisher's Exact Test) and fixed re-sampling (McNemar's Test).

See NCSAT documentation for details.

```{r, echo=FALSE}
library(tidyverse)
library(lubridate)
library(DT)
library(Hmisc)

DT:::DT2BSClass( c("compact", "cell-border"))

dt_opts = list(
  info = FALSE,
  paging = FALSE,
  searching = FALSE,
  scrollX = FALSE,
  scrollY = FALSE,
  ordering = FALSE
  )

```

# Data Evaluation

Data must indicate presence-absence of the species, including survey locations where:

* habitat is appropriate
* survey methods would reasonably detect the species
* detection probablity is similar across time periods (i.e., uses similar methods)

See [Ewing and Gangloff 2016](https://www.researchgate.net/profile/Todd-Ewing/publication/303565271_Using_Changes_in_Naive_Occupancy_to_Detect_Population_Declines_in_Aquatic_Species_Case_Study_Stability_of_Greenhead_Shiner_in_North_Carolina/links/57487e9208ae5f7899b9d6a9/Using-Changes-in-Naive-Occupancy-to-Detect-Population-Declines-in-Aquatic-Species-Case-Study-Stability-of-Greenhead-Shiner-in-North-Carolina.pdf) for more details.

# User Variables
Begin by modifying the following variables for your analysis.

```{r}
setwd(
    paste0(
        "C:\\Users\\skanderson\\OneDrive - State of North Carolina\\",
        "Documents - WRC_Wildlife Management Division\\Wildlife Diversity ",
        "program\\projects\\species_listing\\NCSAT_Process_Guide\\Testing ",
        "NCSAT\\birds"
    )
)
# setwd("~/")

```
```{r, echo=FALSE}
#setup example data table
f_ex <- "./data/example_data_jhall.csv"
data_ex <- read.csv(
  f_ex,
  sep = ",",
  header = TRUE
)

```

### File Location and Format
This file should contain zero-filled data, with one row per survey event in a csv (comma-separated values text file) and contain the following fields:

 * presence/absence (TRUE/FALSE)
 * year
 * location (unique identifier for the survey site)

An example data table would look like this:
```{r, echo=FALSE}
datatable(data_ex[1:5, ], options = dt_opts, width = "75%")
```

Set the year and the presence/absence field (boolean) below:
```{r}
presence_field <- "species_observed"
year <- "year"
location_id <- "locality_id"
```

Set file location relative to working directory:
```{r}
f_zf <- "./data/ebd_sesp_nc_coast_zf_breeding.csv"
# f_zf <- "./data/example_data_jhall.csv"
```

### Generation Length

Set generation length. See NCSAT documentation for calculating.
```{r}
# gen_len <- 4 # in years
gen_len <- "unknown" # unknown generation length
```

### Latest year
Set the most recent year of data to use in the analysis (i.e., the end of the "recent" time period).

Set to "calculate" if you want to use the last year of the data.

```{r}
latest_year <- "calculate"
# Year example:
# latest_year <- 2022
```

### Confidence Level
Set values for running Fisher Exact test:

* alpha: the confidence level of the test, usually 0.90 or 0.95 (i.e., there is only a 10% or 5%, respectively, likelihood the data could occur from a random process)

* z: the z value for the corresponding alpha level

* alt: determines the definition of the null hypothesis to test (signficiant p-values indicate rejection of hull hypothesis)

  * greater = historical/past is less than or equal to recent occupancy

  * less = historical/past is greater than or equal to recent occupancy

  * two-sided = historical/past is the same as recent occupancy

```{r}

ci <- data.frame(
  alpha = 0.9,
  z = 1.645,
  alt = "greater"
)

# note
# z_90 <- 1.645
# z_95 <- 1.96
```

### Analyses to Run
Set as TRUE/FALSE for the analyses to run.

  * Fixed Resampling: appropriate for repeated visits to the same site/location
  * Random Resampling: appropriate for visits to different sites/locations between years

```{r}
  random_resampling <- TRUE
  fixed_resampling <- TRUE
```


```{r, echo=FALSE}
#################################################################
# Set up functions to be used in the script below



################################
# statistical analysis input table
# create a matrix from passed data frame for statistical analysis
make_matrix <- function(tp1, tp2, rownames){
  
  my_matrix <-  matrix(
    c(
      c(tp1$detected, tp2$detected),
      c(tp1$absent, tp2$absent)
      ),
    ncol = 2
  )
  rownames(my_matrix) <- rownames
  colnames(my_matrix) <- c("detected","absent")
  my_matrix
}

################################
# time period statistics
# create a table of time period statistics for the passed list of values
df_calc_stats_cols <- c(
  "tp",
  "method",
  "n",
  "detected",
  "absent",
  "naive_occ",
  "sd",
  "se",
  "lcl",
  "ucl"
)

################################
# create blank tp_stats dataframe
tp_stats <- data.frame(
  matrix(
    nrow = 0,
    ncol = length(df_calc_stats_cols)
  )
)
colnames(tp_stats) = df_calc_stats_cols

df_calc_stats <- function(df, tp, method, sites = c()){
  if (length(sites) > 0) {
    # fixed analysis, return only passed list of sites
    df <- df[df$site %in% sites,]
  }

  det <- sum(df$detected) # number of sites with detection
  n <- nrow(df) # number of records
  abs <- n - det # number of sites without detections
  df_sd <- sd(df$detected) # standard deviation
  df_se <- df_sd / sqrt(n) # standard error
  df_mean <- mean(df$detected) # mean
  results <- data.frame(
    tp = tp,
    method = method,
    n = n,
    detected = det,
    absent = abs,
    naive_occ = df_mean,
    sd = df_sd,
    se = df_se,
    lcl = df_mean - (ci$z * df_se),
    ucl = df_mean + (ci$z * df_se)
  )
  results
}

################################################################
## CONSIDER WRITING FUNCTION TO RUN ANALYSES AND OUTPUT COMPOENENTS
get_comparison <- function(
  tp1,
  tp1_name,
  tp2,
  tp2_name,
  analysis_type,
  compare
  ) {
  # inputs:
  #   tp1 = time period 1 data
  #   tp2 = time period 2 data
  #   analysis_type = "fixed" or "random"
  #   compare = time period list (e.g. c("past", "recent"))
  # outputs:
  #   no_change,
  #   no_change_decline,
  #   change_text,
  #   comp_table_values,
  #   data_table,
  #   test_matrix,
  if (analysis_type == "fixed"){
    # get list of sites surveyed in both time periods
    s <- intersect(tp1$site, tp2$site)
    tp1 <- df_calc_stats(
      tp1,
      tp = tp1_name,
      method = analysis_type,
      sites = s
    )
    tp2 <- df_calc_stats(
      tp2,
      tp = tp2_name,
      method = analysis_type,
      sites = s
    )
  } else {
    tp1 <- df_calc_stats(tp1, tp = tp1_name, method = analysis_type)
    tp2 <- df_calc_stats(tp2, tp = tp2_name, method = analysis_type)
  }


  # format text for tables
  nc <- ((tp2$naive_occ - tp1$naive_occ) / tp1$naive_occ) * 100
  nc <- f_dec(nc, d = 3)
  ncd <- f_dec(-1 * nc, d = 3)
  ct <- paste0(
    "Data suggests ",
    f_dec(nc, d = 2),
    "% change between past and recent naive occupancy values."
  )
  comp_text <- build_compare_text(compare)
  ncd_text <- paste0(ncd, "%")
  tp1_occ_text <- build_occ_text(tp1)
  tp2_occ_text <- build_occ_text(tp2)
  comp_n <- sum(tp1$n, tp2$n)


  # update tp_stats df, if not already in table
  tp_method_vals <- paste0(tp_stats$tp, tp_stats$method)
  if (!(paste0(tp1$tp, tp1$method) %in% tp_method_vals)) {
    tp_stats <<- rbind(
      tp_stats,
      tp1
    )
  }
  if (!(paste0(tp2$tp, tp2$method) %in% tp_method_vals)) {
    tp_stats <<- rbind(
      tp_stats,
      tp2
    )
  }

  # stats matrix
  # format data for stats test
  tm <- make_matrix(
    tp1,
    tp2,
    rownames = compare
  )
  
  # create data table
  dt_data <- tp_stats[
    tp_stats$tp %in% compare & tp_stats$method == analysis_type,
    dt_cols
  ] %>%
  mutate(
    naive_occ = f_dec(naive_occ, d = 3)
  )

  dt <- datatable(
    dt_data,
    options = dt_opts,
    colnames = comparison_colnames
  )

  results <- list(
    no_change = nc,
    no_change_decline = ncd,
    change_text = ct,
    comp_text = comp_text,
    ncd_text = ncd_text,
    tp1_occ_text = tp1_occ_text,
    tp2_occ_text = tp2_occ_text,
    comp_n = comp_n,
    test_matrix = tm,
    data_table = dt
  )

  return(results)

}


################################
# data formatting functions

# shortcut to format numbers with 3 digits
f_dec <- function(v, d = 3) {
  v <- as.numeric(v)
  round(v, digits = d)
}

# shortcut to format p-values to be <0.0001 if super low
f_p <- function(p) {
  p <- as.numeric(p)
  ifelse(
    p < 0.0001,
    "<0.0001",
    as.character(p)
  )
}

# shortcut to format occupancy with Confidence Levels
build_occ_text <- function(tp) {
  # pass the row from the dataframe
  paste0(
    f_dec(tp$naive_occ), 
    " (",
    f_dec(tp$lcl),
    " - ",
    f_dec(tp$ucl),
    ")"
  )
}

# shortcut to format compare text nicely
build_compare_text <- function(tps, pretty = FALSE) {
  if (pretty) {
    paste0(
      str_to_title(tps[1]),
      " v ",
      str_to_title(tps[2])
    )
  } else {
    paste0(
      tps[1],
      "_v_",
      tps[2]
    )

  }
}

################################
# data comparison tables

# column list
dt_cols <- c("tp", "method", "detected", "absent", "n", "naive_occ")

# pretty column names
comparison_colnames <- c(
  "Time Period",
  "Method",
  "Detected",
  "Absent",
  "Total",
  "Naive Occupancy"
)

################################
# final result comparison table
# create blank results comparison table
results_comp <- data.frame(
  comparison = character(),
  method = character(),
  pct_decline = character(),
  initial_occupancy = character(),
  recent_occupancy = character(),
  n = integer(),
  p = character()
)

```


# Set Up Data For Statistical Analysis

### Load the Dataset
```{r}
data_zf <- read.csv(
  f_zf,
  sep = ",",
  header = TRUE
)
```

### Calculate Time Periods
See documentation for calculating generation length.

```{r}
# determine three generation length
if (!is.numeric(gen_len)) {
  three_gen <- 10
  gen_len <- f_dec(three_gen / 3, d = 0)
} else {
  three_gen <- 3 * gen_len
}
# determine latest year
if (!is.numeric(latest_year)){
  latest_year <- max(data_zf[[year]])
} else {
  latest_year <- as.numeric(latest_year)
}

# recent will always be a time period
recent_start <- latest_year - (0.5 * three_gen)
recent_end <- latest_year

#past
past_start <- latest_year - three_gen
past_end <- recent_start - 1

# historical
oldest_year <- min(data_zf[[year]])
hist_start <- oldest_year
hist_end <- recent_start - 1

# Time periods

tp <- data.frame(
  c(recent_start, past_start, hist_start),
  c(recent_end, past_end, hist_end)
)
rownames(tp) <- c("Recent", "Past", "Historical")
colnames(tp) <- c("Start", "End")
```

### Time Period Begin and End Years

```{r echo=FALSE}
datatable(tp, options = dt_opts, width = "50%")
```
```{r, results='asis',echo=FALSE}
# place a statement of generation times
cat(
  paste0(
    gen_len,
    " Year Generation Time",
    "  --  ",
    three_gen,
    " Year Three Generation Time"
  )
)
```

### Reformat Data
Prepare data for analysis.

```{r}

# list of fields to add to the data set
fet_fields <- c(
  "site", # unique id for the site surveyed
  "recent", # boolean for recent time period
  "past", # boolean for past time period
  "historical", # boolean for historical time period
  "detected" # 1 if detected, 0 if not
)

# add fields to data frame, return data frame with only required fields
data_zf <- data_zf %>%
  mutate(
    detected = ifelse(
      get(presence_field),
      1,
      0
    ),
    recent = (!!as.name(year) >= recent_start & !!as.name(year) <= recent_end),
    past = (!!as.name(year) >= past_start & !!as.name(year) <= past_end),
    historical = (!!as.name(year) >= hist_start & !!as.name(year) <= hist_end),
    site = (!!as.name(location_id))
  ) %>%
  select(all_of(fet_fields))

# get encounter histories and stats for each time period
# build table of stats for each time period

########################################################
### NEEDS ATTENTION
### For now, shrinks data to only have one row per
### site-time period combination. Uses the maximum value
### for detection. This ignores different sampling effort
### during each time period (i.e., more visits to a site
### increases chances of detection).
### See group_by and summarise stages below

### Adds site_visits column with count of number of surveys
### during the time period.

### To Do: confirm original methodology for use of
### McNemar's Test - modify code to match

########################################################

recent <- data_zf  %>%
  filter(recent) %>%
  group_by(site) %>%
  summarise(
    recent = max(recent),
    past = max(past),
    historical = max(historical),
    detected = max(detected),
    site_visits = n()
  )

past <- data_zf  %>%
  filter((past)) %>%
  group_by(site) %>%
  summarise(
    recent = max(recent),
    past = max(past),
    historical = max(historical),
    detected = max(detected),
    site_visits = n()
  )

historical <- data_zf  %>%
  filter((historical)) %>%
  group_by(site) %>%
  summarise(
    recent = max(recent),
    past = max(past),
    historical = max(historical),
    detected = max(detected),
    site_visits = n()
  )
```

# Past vs. Recent

## Random Re-Sampling
Appropriate where diffeent sites are sampled across time periods.

```{r}
# function to get comparison information
pr_random_comp <- get_comparison(
  tp1 = past,
  tp1_name = "past",
  tp2 = recent,
  tp2_name = "recent",
  analysis_type = "random",
  compare = c("past", "recent")
)

```

### Frequency Table

```{r}

pr_random_comp$data_table
```
```{r, results='asis',echo=FALSE}
cat(pr_random_comp$change_text)

```

### Results

```{r eval = random_resampling}
# get data for past v recent random comparison

pr_random_results <- fisher.test(
  pr_random_comp$test_matrix,
  conf.level = ci$alpha,
  alternative = ci$alt
)

#add these results to the comparison table
results_comp[nrow(results_comp) + 1, ] <-
  c(
    pr_random_comp$comp_text,
    "Random",
    pr_random_comp$ncd_text,
    pr_random_comp$tp1_occ_text,
    pr_random_comp$tp2_occ_text,
    pr_random_comp$comp_n,
    format(
      c( f_p( pr_random_results$p )),
      digits = 4
    )
  )


pr_random_results

```


## Fixed Re-Sampling
Appropriate when the same sites are sampled across time periods.

```{r}

# function to get comparison information
pr_fixed_comp <- get_comparison(
  tp1 = past,
  tp1_name = "past",
  tp2 = recent,
  tp2_name = "recent",
  analysis_type = "fixed",
  compare = c("past", "recent")
)

```

### Frequency Table

```{r}
pr_fixed_comp$data_table
```
```{r, results='asis',echo=FALSE}

cat(pr_fixed_comp$change_text)

```

### Results
```{r}
if (fixed_resampling){
  pr_fixed_results <- mcnemar.test(
    pr_fixed_comp$test_matrix
  )

  #add these results to the comparison table
  
results_comp[nrow(results_comp) + 1, ] <-
  c(
    pr_fixed_comp$comp_text,
    "Fixed",
    pr_fixed_comp$ncd_text,
    pr_fixed_comp$tp1_occ_text,
    pr_fixed_comp$tp2_occ_text,
    pr_fixed_comp$comp_n,
    format(
      c( f_p( pr_random_results$p.value)),
      digits = 4
    )
  )

  pr_fixed_results
}
```

# Historical vs. Recent

## Random Re-Sampling
Appropriate when different sites are sampled across time periods.

```{r}
# function to get comparison information
hr_random_comp <- get_comparison(
  tp1 = historical,
  tp1_name = "historical",
  tp2 = recent,
  tp2_name = "recent",
  analysis_type = "random",
  compare = c("historical", "recent")
)

```

### Frequency Table

```{r}
# put data in table for formatting output
hr_random_comp$data_table
```
```{r, results='asis',echo=FALSE}
# read out comparison results
cat(hr_random_comp$change_text)
```

### Results
```{r}
if (random_resampling) {

  hr_random_results <- fisher.test(
    hr_random_comp$test_matrix,
    conf.level = ci$alpha,
    alternative=ci$alt
  )

  # add these results to the comparison table
  results_comp[nrow(results_comp) + 1, ] <-
    c(
      hr_random_comp$comp_text,
      "Random",
      hr_random_comp$ncd_text,
      hr_random_comp$tp1_occ_text,
      hr_random_comp$tp2_occ_text,
      hr_random_comp$comp_n,
      format(
        c( f_p(hr_random_results$p )),
        digits = 4
      )
    )

  hr_random_results
}
```
## Fixed Re-Sampling

Appropriate when the same sites are sampled across time periods.

```{r}
# function to get comparison information
hr_fixed_comp <- get_comparison(
  tp1 = historical,
  tp1_name = "historical",
  tp2 = recent,
  tp2_name = "recent",
  analysis_type = "fixed",
  compare = c("historical", "recent")
)

```

### Frequency Table

```{r}
hr_fixed_comp$data_table
```
```{r, results='asis',echo=FALSE}
cat(hr_fixed_comp$change_text)
```

### Results
```{r}
if (fixed_resampling){

  hr_fixed_results <- mcnemar.test(
    hr_fixed_comp$test_matrix
  )

  # add these results to the comparison table
  results_comp[nrow(results_comp) + 1,] = 
    c(
      hr_fixed_comp$comp_text,
      "Fixed",
      hr_fixed_comp$ncd_text,
      hr_fixed_comp$tp1_occ_text,
      hr_fixed_comp$tp2_occ_text,
      hr_fixed_comp$comp_n,
      format(
        c( f_p(hr_fixed_results$p.value)),
        digits = 4
      )
    )

  hr_fixed_results
}
```

# Results Comparison

NCSAT Criterion A Data Table

```{r}

result_cols <- c(
  "",
  "Comparison",
  "Method",
  "% Decline",
  "Initial Occupancy (+/-90% CL)",
  "Recent Occupancy (+/- 90% CL)",
  "N",
  "p")

datatable(results_comp, colnames = result_cols, options = dt_opts)
```




